Unifying intraspecific and interspecific variation in tropical tree mortality

### 2016-06-15 ms/ms.Rnw
Partly reorganise manuscript
Move sections of methods to end of introduction

Will need to re-evaluate the length, but this improves readability for now.

\begin{abstract}
Add abstract here
\end{abstract}

@@

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{P}lant mortality is often modelled as a function of recent growth because it reflects the product of many biotic and abiotic factors on a plants overall carbon budget and vigour \cite{Kobe:1995tw, Hawkes:2000ib, Keane:2001db, Wyckoff:2002ul}. As such, growth rates are considered a highly integrated individual-based predictor of plant mortality \cite{Hawkes:2000ib}. Within a species, the assumption is that fast growing individuals (i.e. those that have larger carbon budgets) are more able to buffer stress and therefore have higher survivorship relative to slow growing (i.e. low carbon budget) individuals \cite{Hawkes:2000ib, Keane:2001db}. Empirical studies examining this growth\textendash mortality relationship have broadly supported this assumption, with many highlighting non-linear declines in mortality with increased growth \cite[e.g.][]{Kobe:1997vy, Wyckoff:2002ul,Vieilledent:2010fv}. Consequently, most dynamic vegetation models, including tree gap models \cite{Keane:2001db}, individual-based forest simulators \cite[e.g. SORTIE][]{Pacala:1996vm} and Dynamic Global Vegetation Models \cite[DGVMs; ][]{Woodward:2004ft} use some measure of growth coupled with a stochastic constant (i.e. baseline hazard or minimum chance of death) as the fundamental determinants of plant mortality \cite{Hawkes:2000ib, McDowell:2011dr}.

@@

\begin{figure}[ht]
\centering
\includegraphics{figures/fig1.pdf}
\caption{Outline of methodology. A) Past growth used to predict future mortality. B) Three functional forms of plant mortality. C) Cross validation proceedure}
\label{fig1}
\end{figure}

In this study, we reconcile both intraspecific and interspecific theory by developing a unfied model that incorporates both individual growth and the species level trait, wood density. We extend existing individual-based approaches \cite[e.g.][]{Kobe:1995tw} by simultaneously estimating both baseline hazards and the growth-mortality relationship within a single model (hereafter referred to as growth-dependent hazard). We then compare the predictive performance of this model to other traditionally used mortality models. We parameterize these models using 15-years of individual growth and mortality data from 203 tropical rainforest species encompassing \Sexpr{n_inds} individuals. Using these models, coupled with 10-fold cross validation, we ask several questions: 1) Does simultaneously estimating baseline and growth-dependent hazards improve predictive capacity relative to baseline only or growth-dependent only models? 2) is DBH growth or basal area growth more predictive of tropical tree mortality and does it's predictive capacity vary with model type? 3) Does allowing both hazards to vary by species improve predictive capacity? 4) Does wood density predominately affect baseline hazards, or does it affect growth-dependent hazards or both?

@@

\subsection*{Mortality analysis}

In survival analysis, the probability that an individual dies between $t_2$ and $t_3$, $p_{t_2\rightarrow t_3}$ is defined by:
\begin{equation} \label{eq:mortality_model}
p_{i,t_2\rightarrow t_3} = 1 - \exp\left(- \int_{t_2}^{t_3} \lambda_i(t) {\rm d}t\right).
\end{equation}
where, $\lambda$ is the instantaneous hazard rate of death per unit time \cite{Kleinbaum:2012tn} and $\int_{t_2}^{t_3} \lambda_i(t) {\rm d}t$ is the cumulative hazard between $t_2$ and $t_3$.

Here we examine the performance of three functional forms of $\lambda$ to predict observed patterns of tropical tree mortality. The simplest model (eq. \ref{base_haz}), assumes a single baseline hazard, which is constant irrespective of a plant's growth rate. The second model (eq. \ref{growth_haz} describes a growth-dependent hazard which assumes plants can only die due to a growth related cause. In this model, we make an explicit assumption that the growth-mortality relationship is captured by a negative exponential curve whereby the probability of death is high at low growth rates and as growth increases the probability of dying decays at an exponential rate until an asymptote at zero is reached. This negative exponential relationship has been used in other studies to model growth-dependent mortality \cite[e.g.][]{Kobe:1995tw} and broadly reflects patterns observed in non-parametric growth-mortality models \cite{Wyckoff:2000un, Kunstler:2005bn, Vieilledent:2010fv}. The third model (eq. \ref{add_haz}, and potentially the most biologically realistic, is the summation of the two previous models. Specifically, this model allows $\lambda$ to decrease exponentially with increasing growth, but, unlike a standard negative exponential, will asymptote at some baseline hazard instead of zero. The advantage of this model is that it effectively accounts for two independent causes of mortality: death due to growth and death due to chance (i.e. stochastic events such as wind throw and fire). The parameters of eq. \ref{add_haz} are also biologically interpretable: $\alpha$ represents the intercept and defines the instantaneous mortality hazard $\lambda$ at low annual growth rate. $\beta$ represents the exponential decay rate or the sensitivity of hazard rates to change with growth rate; and $\gamma$, is the asymptote and defines the baseline hazard. $\alpha$ and $\beta$ are intrinsically linked to a plants growth rate, and as such are growth-dependent parameters. By contrast, $\gamma$ is a growth-independent parameter that accounts for stochastic events (e.g. windfall, fire) that can kill a plant irrespective of its growth rate.
\begin{align}
\lambda_i &= \overbrace{\gamma}^{\text{baseline hazard}} \label{base_haz}\\
\lambda_i &= \overbrace{\alpha \, \exp\left(-\beta \, X_i \right)}^{\text{Growth-dependent hazard}} \label{growth_haz}\\
\lambda_i &= \overbrace{\alpha \, \exp\left(-\beta \, X_i \right)}^{\text{Growth-dependent hazard}} +
\overbrace{\gamma}^{\text{Baseline hazard}} \label{add_haz}
\end{align}

The central aim of this study is to identify the most parsimonious combination of $\lambda$, growth measure (diameter increment growth or area growth) and wood density that best predicts tropical tree mortality. To do this, we used K-fold cross-validation and four iterative stages of model fitting and comparison. Stage 1 involved comparing predictive performance of the three $\lambda$ models described in equations \ref{base_haz} \textendash \ref{add_haz} using either annual diameter increment or area growth rates. In this stage each model was fitted without allowing parameters to vary by species. This allowed us to determine: 1) whether predictive performance was substantially improved with increasing functional complexity; and 2) whether predictive performance varied by growth measure. Following this, we used the best model of stage 1 but allowed its parameters to vary by species. This allowed us to determine how variable each parameter was across species and whether accounting for species differences significantly improved predictive performance. Because species responses varied considerably, we undertook a third stage of model fitting and selection that explicitly examined whether wood density (a species level trait) explained some of this interspecific variability. This involved using the best model observed in stage 1 and adding wood density effects to each parameter. Because we were interested in obtaining the most parsimonious model of tropical tree mortality, we fitted the model under all combinations of wood density effects. This allowed us to determine the most parsimonious combination of wood density effects and whether it most strongly influenced growth-dependent or growth-independent mortality hazards. Following this we then fitted the best wood density combination model with species random effects to the entire dataset. This model was then used to examine mortality curves across wood density gradients. In order to determine the proportion of interspecific variation explained by wood density on each parameter it occurred on we reran the final model form with just random effects on parameters and calculated the percent reduction in unexplained species variation.

\subsection*{Model fitting}

Our observations are for survival outcomes $S_{i, t_2\rightarrow t_3}$ (0 = alive, 1 = died) for each individual $i$ between censuses at time $t_2$ and $t_3$. We model these observations as random realizations from a Bernoulli distribution
\begin{equation}
S_{i,t_2\rightarrow t_3} \sim \textrm{Bernoulli}(p_{i,t_2\rightarrow t_3}),
\end{equation}
where $p_{i,t_2\rightarrow t_3}$ is the probability individual $i$ dies in the interval $t_2$--$t_3$. Using standard survival analysis, this probability is given by
\begin{equation}
p_{i,t_2\rightarrow t_3} = 1 - \exp\left(- \int_{t_2}^{t_3} \lambda_i(t) {\rm d}t\right).
\end{equation}
As we only have a point estimate of $\lambda_i(t)$ at $t=t_2$, the cumulative hazard from  $t_2$--$t_3$ is approximated as
\begin{equation}
  \int_{t_2}^{t_3} \lambda_i(t) {\rm d}t \approx \lambda_i(t_2) (t_3-t_2).
\end{equation}

@@

We standardized wood density via the transformation: $\frac{\rho}{\rho_c}$), where $\rho_c=0.6$ g cm${^-3}$. This operation is equivalent to centering a log-transformed $\rho$ around $\rho_c$ and enables mean hyper-parameters  to be interpreted as the value for a species with average wood density of 0.6 cm${^-3}$. Similarly, we also centered $X_i$ at the lower 5\% quantile for both diameter increment and area growth (0.172 and 0.338, respectively), meaning $\alpha_s$ should be interpreted as the hazard rate when growth rate was very low.

@@

\subsection*{Model evaluation}
We examined predictive accuracy of each model by using K-fold cross-validation. This involved randomly partitioning the data into ten folds of \Sexpr{n_heldout} observations, while ensuring that each K-fold had at least one observation per species. We then fitted models to nine folds and predicted to the tenth fold. This was repeated until each fold had been used nine times for model fitting, and once as a held-out dataset for predicting. Cross-validation is a typical approach to model evaluation, and assesses a model’s accuracy when extrapolating to data not used in model calibration, thereby indicating the model’s predictive accuracy \cite{Stone:1997tg}.

Predictive accuracy was quantified by estimating the average log loss across K-folds for held-out data, $\bar{\mathcal{L}}$ (Fig. \ref{fig1}). Logarithmic loss -- commonly known as log loss, and indicated by $\mathcal{L}$-- measures the accuracy of a model by penalizing incorrect predictions based on how wrong the predicted probability is from the observed outcome (Fig ??). By minimizing the $\bar{\mathcal{L}}$, we maximize predictive skill. The average log loss across all individuals for the $k$th fold of held-out data, $\mathcal{L}_k$, is given by
\begin{equation} \label{eq:logloss}
\mathcal{L}_k = - \frac1{N_k}\,\sum_{i=1}^{N_k}\big(S_i\,\log(p_{i,t_2\rightarrow t_3}) + ( 1 - S_i)\,\log(1 - p_{i,t_2\rightarrow t_3})\big),
\end{equation}
where $N_k$ refers to the number of observations in fold.

>>>>

\begin{abstract}
Individual death is a key process driving population dynamics, nutrient cycling, and evolution within plant communities. While past research has identified factors across a variety of scales influencing tree mortality -- including an individual's past growth, species-level functional traits, and temporal plot-scale effects; these different factors are yet to be integrated within a predictive framework. Here we apply classic techniques from survival analysis within a hierarchical Bayesian and cross-validation framework to derive instantaneous mortality functions for tropical rain forest trees at Barro Colorado Island (BCI) Panama, integrating individual-level, species-level, and plot-level effects. In reaching this point, we show how mortality models including separate growth-dependent and growth-independent components, using stem diameter growth rather than basal-area growth, and attributing the effects of wood density to growth-independent death outperform alternative formulations. Moreover, we show that the effects of wood density -- a prominent trait known to influence mortality -- capture only a portion of total species-level differences; and that the year-to-year variation in mortality was large.  Our analysis also suggests that growth-dependent processes -- and thus competition -- predominate as the biggest driver of mortality at BCI. Combined, this study provides a framework for predicting individual-level mortality in highly diverse tropical forests, but also highlights how little we now about the causes of species-level and temporal plot-scale effects, needed to effectively predict tree mortality.
\end{abstract}

@@

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.

\dropcap{P}lant mortality is often modelled as a function of recent growth because it reflects the product of many biotic and abiotic factors on a plants overall carbon budget and vigour \cite{Kobe:1995tw, Hawkes:2000ib, Keane:2001db, Wyckoff:2002ul}. As such, growth rates are considered a highly integrated individual-based predictor of plant mortality \cite{Hawkes:2000ib}. Within a species, the assumption is that fast growing individuals (i.e. those that have larger carbon budgets) are more able to buffer stress and therefore have higher survivorship relative to slow growing (i.e. low carbon budget) individuals \cite{Hawkes:2000ib, Keane:2001db}. Empirical studies examining this growth\textendash mortality relationship have broadly supported this assumption, with many highlighting non-linear declines in mortality with increased growth \cite[e.g.][]{Kobe:1997vy, Wyckoff:2002ul,Vieilledent:2010fv}. Consequently, most dynamic vegetation models, including tree gap models \cite{Keane:2001db}, individual-based forest simulators \cite[e.g. SORTIE][]{Pacala:1996vm} and Dynamic Global Vegetation Models \cite[DGVMs; ][]{Woodward:2004ft} use some measure of growth coupled with a stochastic constant (i.e. baseline hazard or minimum chance of death) as the fundamental determinants of plant mortality \cite{Hawkes:2000ib, McDowell:2011dr}.

@@

\begin{figure}[ht]
\centering
\includegraphics{figures/fig1.pdf}
\caption{Outline of methodology. A) Our data consist of repeat measures on stem diameter ($D$) and status ($S$, alive or dead) for individual tress at specific census dates ($t_1$, $t_2$, $t_3$). B) We consider three alternative mortality functions, combining growth-independent and growth-dependent effects. For each model form, we consider two alternative predictors of growth, $X$ (basal area and stem diameter growth), as well as allowing for species-level effects on the parameters $\alpha$, $\beta$ and $\gamma$. C) The predictive skill of alternative models is evaluated via 10-fold cross validation. The entire dataset is split into 10 folds ($F_1$, ..., $F_{10}$). We then refit each alternative model 10 times ($M_1$, ..., $M_{10}$), using different combinations of testing and training data. We then average the log loss $\mathcal{L_i}$ for each of these 10 model fits, to give an overall average log loss for each model $\bar{\mathcal{L}}$.}
\label{fig1}
\end{figure}

In this study, we reconcile both intraspecific and interspecific theory by developing a unified model that incorporates both individual growth and the species level trait, wood density. We extend existing individual-based approaches \cite[e.g.][]{Kobe:1995tw} by simultaneously estimating both baseline hazards and the growth-mortality relationship within a single model (hereafter referred to as growth-dependent hazard). We then compare the predictive performance of this model to other traditionally used mortality models. We parameterize these models using 15-years of individual growth and mortality data from 203 tropical rainforest species encompassing \Sexpr{n_inds} individuals. Using these models, coupled with 10-fold cross validation, we ask several questions: 1) Does simultaneously estimating baseline and growth-dependent hazards improve predictive capacity relative to baseline only or growth-dependent only models? 2) is DBH growth or basal area growth more predictive of tropical tree mortality and does it's predictive capacity vary with model type? 3) Does allowing both hazards to vary by species improve predictive capacity? 4) Does wood density predominately affect baseline hazards, or does it affect growth-dependent hazards or both?

\section*{Framework for modelling mortality}

In classic survival analysis, the probability that an individual dies between $t_2$ and $t_3$, $p_{t_2\rightarrow t_3}$ is defined by
\begin{equation} \label{eq:mortality_model}
p_{i,t_2\rightarrow t_3} = 1 - \exp\left(- \int_{t_2}^{t_3} \lambda_i(t) {\rm d}t\right),
\end{equation}
where $\lambda_i(t)$ is the instantaneous hazard (mortality) rate of death per unit time \cite{Kleinbaum:2012tn} and $\int_{t_2}^{t_3} \lambda_i(t) {\rm d}t$ is the cumulative hazard between $t_2$ and $t_3$. The challenge then is to identify suitable forms of the instantaneous hazard (mortality) rate function.

We examine the performance of three functional forms of $\lambda_i(t)$ to predict observed patterns of tropical tree mortality:
\begin{align}
\lambda_i &= \overbrace{\gamma}^{\text{baseline hazard}} \label{base_haz}\\
\lambda_i &= \overbrace{\alpha \, \exp\left(-\beta \, X_i \right)}^{\text{Growth-dependent hazard}} \label{growth_haz}\\
\lambda_i &= \overbrace{\alpha \, \exp\left(-\beta \, X_i \right)}^{\text{Growth-dependent hazard}} +
\overbrace{\gamma.}^{\text{Baseline hazard}} \label{add_haz}
\end{align}
The simplest model (eq. \ref{base_haz}), assumes a single baseline hazard, which is constant irrespective of a plant's growth rate. The second model (eq. \ref{growth_haz}) describes a growth-dependent hazard which assumes plants can only die due to a growth related cause. In this model, we make an explicit assumption that the growth-mortality relationship is captured by a negative exponential curve whereby the probability of death is high at low growth rates and as growth increases the probability of dying decays at an exponential rate until an asymptote at zero is reached. This negative exponential relationship has been used in other studies to model growth-dependent mortality \cite[e.g.][]{Kobe:1995tw} and broadly reflects patterns observed in non-parametric growth-mortality models \cite{Wyckoff:2000un, Kunstler:2005bn, Vieilledent:2010fv}. The third model (eq. \ref{add_haz}) -- potentially the most biologically realistic -- is the summation of the two previous models. Specifically, this model allows $\lambda$ to decrease exponentially with increasing growth, but, unlike a standard negative exponential, will asymptote at some baseline hazard instead of zero. The advantage of this model is that it effectively accounts for two independent causes of mortality: death due to growth and death due to chance (i.e. stochastic events such as wind throw and fire). The parameters of eq. \ref{add_haz} are also biologically interpretable: $\alpha$ represents the intercept and defines the instantaneous mortality hazard $\lambda$ at low annual growth rate. $\beta$ represents the exponential decay rate or the sensitivity of hazard rates to change with growth rate; and $\gamma$, is the asymptote and defines the baseline hazard. $\alpha$ and $\beta$ are intrinsically linked to a plants growth rate, and as such are growth-dependent parameters. By contrast, $\gamma$ is a growth-independent parameter that accounts for stochastic events (e.g. windfall, fire) that can kill a plant irrespective of its growth rate.

The central aim of this study is to identify the most parsimonious combination of $\lambda$, growth measure (diameter increment growth or area growth) and wood density that best predicts tropical tree mortality. To do this, we used K-fold cross-validation and four iterative stages of model fitting and comparison. Stage 1 involved comparing predictive performance of the three $\lambda$ models described in equations \ref{base_haz} \textendash \ref{add_haz} using either annual diameter increment or area growth rates. In this stage each model was fitted without allowing parameters to vary by species. This allowed us to determine: 1) whether predictive performance was substantially improved with increasing functional complexity; and 2) whether predictive performance varied by growth measure. Following this, we used the best model of stage 1 but allowed its parameters to vary by species. This allowed us to determine how variable each parameter was across species and whether accounting for species differences significantly improved predictive performance. Because species responses varied considerably, we undertook a third stage of model fitting and selection that explicitly examined whether wood density (a species level trait) explained some of this interspecific variability. This involved using the best model observed in stage 1 and adding wood density effects to each parameter. Finally we then fitted the best wood density combination model with species random effects to the entire dataset. This model was then used to examine mortality curves across wood density gradients. In order to determine the proportion of interspecific variation explained by wood density on each parameter it occurred on we reran the final model form with just random effects on parameters and calculated the percent reduction in unexplained species variation.

At each stage, we evaluated the predictive accuracy of alternative models or predictor variables using K-fold cross-validation (Fig. \ref{fig1}). This involves randomly partitioning the data into ten folds of \Sexpr{n_heldout} observations, while ensuring that each K-fold had at least one observation per species. We then fitted models to nine folds and predicted to the tenth fold. This was repeated until each fold had been used nine times for model fitting, and once as a held-out dataset for predicting. Cross-validation is a typical approach to model evaluation, and assesses a model’s accuracy when extrapolating to data not used in model calibration, thereby indicating the model’s predictive accuracy \cite{Hooten:2015, Stone:1997tg}.

@@

\subsection*{Model fitting}
Our observations are for survival outcomes $S_{i, t_2\rightarrow t_3}$ (0 = alive, 1 = died) for each individual $i$ between censuses at time $t_2$ and $t_3$. We model these observations as random realizations from a Bernoulli distribution
\begin{equation}
S_{i,t_2\rightarrow t_3} \sim \textrm{Bernoulli}(p_{i,t_2\rightarrow t_3}),
\end{equation}
where $p_{i,t_2\rightarrow t_3}$ is the probability individual $i$ dies in the interval $t_2$--$t_3$ *(eq. \ref{eq:mortality_model}). As we only have a point estimate of $\lambda_i(t)$ at $t=t_2$, the cumulative hazard from  $t_2$--$t_3$ is approximated as
\begin{equation}
  \int_{t_2}^{t_3} \lambda_i(t) {\rm d}t \approx \lambda_i(t_2) (t_3-t_2).
\end{equation}

@@

We standardized wood density via the transformation: $\frac{\rho}{\rho_c}$, where $\rho_c=0.6$ g cm${^-3}$. This operation is equivalent to centering a log-transformed $\rho$ around $\rho_c$ and enables mean hyper-parameters  to be interpreted as the value for a species with average wood density of 0.6 cm${^-3}$. Similarly, we also centered $X_i$ at the lower 5\% quantile for both diameter increment and area growth (0.172 and 0.338, respectively), meaning $\alpha_s$ should be interpreted as the hazard rate when growth rate was very low.

@@

\subsection*{Model evaluation}
Predictive accuracy was quantified by estimating the average log loss across K-folds for held-out data, $\bar{\mathcal{L}}$ (Fig. \ref{fig1}). Logarithmic loss -- commonly known as log loss, and indicated by $\mathcal{L}$-- measures the accuracy of a model by penalizing incorrect predictions based on how wrong the predicted probability is from the observed outcome (Fig ??). By minimizing the $\bar{\mathcal{L}}$, we maximize predictive skill. The average log loss across all individuals for the $k$th fold of held-out data, $\mathcal{L}_k$, is given by
\begin{equation} \label{eq:logloss}
\mathcal{L}_k = - \frac1{N_k}\,\sum_{i=1}^{N_k}\big(S_i\,\log(p_{i,t_2\rightarrow t_3}) + ( 1 - S_i)\,\log(1 - p_{i,t_2\rightarrow t_3})\big),
\end{equation}
where $N_k$ refers to the number of observations in fold.
